A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2016-08-06 imaging_demo/ms.tex
Make changes in notation in mathematical framework
From 8448190 (mnras):
====================

(*) Make changes in notation in section 2 (mathematical framework)

Electric fields from astrophysical sources, $E(\hat{\mathbf{s}})$, in the sky
coordinate system denoted by sine-projected unit vector $\hat{\mathbf{s}}$, 
propagate towards the observer as:
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \int E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}},
\end{align}
where, $\mathbf{r}$ denotes the observer's location (measured in wavelengths 
relative to some arbitrary origin) and $\widetilde{E}
(\mathbf{r})$ is the propagated electric field. Thus the propagated electric
field is a linear superposition of the electric fields emanating from
astronomical sources with appropriate complex phases. It can also be described
as a Fourier transform of the electric fields in the sky coordinates.

An antenna, $a$, measures a phased sum of these propagated electric fields over
its effective collecting area with an additive receiver noise:
\begin{align}\label{eqn:measured-E-field}
  \widetilde{E}_a &= \int \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a)\,\widetilde{E}(\mathbf{r})\,\dif^2\mathbf{r} + \widetilde{n}_a \\
                  &= \int \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a) \left[ \int E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} \right] \dif^2\mathbf{r} + \widetilde{n}_a \\
                  &= \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a
\end{align}
where, $\widetilde{W}_a(\mathbf{r})$ is the aperture electric field illumination
pattern of the antenna and its Fourier transform, $W_a(\hat{\mathbf{s}})$, is the
directional antenna voltage response.

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p &= \left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		&= \left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  &= \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging and $\star$ denotes complex conjugation. This expression can be further 
simplified with the sky brightness, 
$I(\hat{\mathbf{s}})= \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) 
\right\rangle_t  \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the 
antenna pair sky power response function (or the directional antenna power pattern), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the power 
response, and uncorrelated noise terms which we group into $\widetilde{n}_p$.
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where, the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B_p(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  \widetilde{V}_p &= \int \widetilde{B}(\mathbf{u}^\prime-\mathbf{u}) \left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right]\dif^2\mathbf{u} + n_p,
\end{align}
where, $\widetilde{B}(\mathbf{u})$ denotes the $uv$-space antenna power response
obtained by a Fourier transform of $B(\hat{\mathbf{s}})$. Effectively, the
multiplication in image space by $B(\hat{\mathbf{s}})$ has been replaced by a
convolution with $\widetilde{B}(\mathbf{u})$ in $uv$-space. This is the software
holographic equivalent of a traditional FX correlator output.

Hereafter, we adopt the matrix notation of \citet{mor11}, where vectors are 
represented with single coordinates, and matrices are represented by two 
coordinates denoting the spaces the operator transforms between. 
In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\mathbf{I}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\mathbf{I}(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{\mathbf{E}}(\mathbf{a})\,\widetilde{\mathbf{E}}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

Using the optimal map-making formalism \citep{teg97a,teg97b}, a software
holography image is formed using \citep{mor09}:
\begin{align}
  \mathbf{I}^\prime(\hat{\mathbf{s}}) &= \mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{u})\,\widetilde{\mathbf{B}}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v},\mathbf{v})\,\mathbf{m}(\mathbf{v}) \label{eqn:dirty-image-FX}
\end{align}
where the measured visibilities are weighted by the inverse of the system noise,
followed by a gridding process using the holographic antenna power response as 
the gridding kernel, followed by a Fourier transform to create an image
$\mathbf{I}^\prime(\hat{\mathbf{s}})$. This is the optimal estimate of the true 
image $\mathbf{I}(\hat{\mathbf{s}})$ given the visibility measurements.

The intermediate step of gridding with the antenna power response can be 
expressed as a convolution of a data vector generated by gridding the electric 
fields directly with the antenna illumination pattern.
\begin{align}
\widetilde{\mathbf{B}}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v}) &= \left\langle \left[\widetilde{\mathbf{W}}^\textrm{T}_a(\mathbf{r},\mathbf{a})\, \widetilde{\mathbf{N}}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}(\mathbf{a})\right]\right. \nonumber\\ 
&\quad\ast\left.\left[\widetilde{\mathbf{W}}_a(\mathbf{r},\mathbf{a})\, \mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}^\star(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

We can then use the multiplication-convolution theorem to move the convolution in 
Equation \ref{eqn:e-field-conv} to a square after the Fourier transform in 
Equation 
\ref{eqn:dirty-image-FX}.
\begin{align}
  \mathbf{I}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\widetilde{\mathbf{W}}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}^{-1}(\mathbf{a},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as
that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna
electric fields are weighted by the antenna noise, weighted and gridded by the
antenna aperture kernel, Fourier transformed and finally squared to obtain the
same image estimate that would have been obtained using equation~
\ref{eqn:dirty-image-FX}.

@@

% \subsection{Data Flow}

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain an image cube for every time 
stamp. These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).

\begin{figure}
  \includegraphics[width=\columnwidth]{figure1}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then 
    Fourier transformed by the F-engine to produce electric field spectra 
    $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$. They are calibrated 
    and gridded. The gridded electric fields $\widetilde{E}_\textrm{g}(f)$ from 
    each time series are imaged to produce images 
    $I_1(f)\ldots I_\textrm{M}(f)$. These images are time-averaged to obtain the 
    final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility-based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $\widetilde{V}_\textrm{p}(f)$ ($\mathbf{m}(\mathbf{v})$ 
in matrix notation). They are calibrated and time-averaged to obtain 
$\langle \widetilde{V}_\textrm{p}(f)\rangle$ which are then gridded and imaged to 
obtain the image $I(f)$. The $I(f)$ obtained from both techniques are 
theoretically identical as explained in \S\ref{sec:math}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure2}
  % \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility-based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $\widetilde{V}_\textrm{p}(f)$ which are calibrated and 
    time-averaged. Then they are gridded to obtain the gridded visibilities 
    $\widetilde{V}_\textrm{g}(f)$ which are then Fourier transformed to obtain 
    the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

The gridding kernel in the simplest case is given by the antenna aperture
illumination function, $\widetilde{B}(\mathbf{r}-\mathbf{r}_a)$, which can
be specified either by a functional form or as a table of values against 
locations around the antennas. A nearest neighbor mapping from all antenna 
footprints to grid locations is created using an efficient k-d tree algorithm 
\citep{man99}. There is no restriction here that the aperture illumination 
function has to be identical across antennas.

@@

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the 
individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved by
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process using the sparse matrices module in Python SciPy package. 
Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra,
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}).
\end{align}

@@

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centreed at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility-based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration time-scale and does not add significant cost to the full operation.

@@

Here we demonstrate our software using narrow band data from the LWA station 
in New Mexico. This data is in LWA narrow-band transient buffer (TBN) format 
from 255 antennas within roughly a diameter of 100~m. The data is centreed at a 
frequency of 74.03~MHz, with a sample rate (equal to the bandwidth) of 100~kHz 
with 512 complex time samples per antenna in a A/D writeout time-scale of 
5.12~ms, a frequency resolution of 195.3125~Hz and dual polarization. There are 
391 such writeouts (or time stamps; each contains 512 time samples at 100~kHz 
sampling) yielding a total duration of 2~s.

@@

Besides its core functions of spatial Fourier transform and squaring, it can 
stack, accumulate and average images, and optionally remove the antenna
auto-correlations centreed around the zero-spacing pixel in the $uv$ plane. 
It also handles all four cross-polarization products. Currently, it supports 
writing data out in standard FITS format.

>>>>

Electric fields from astrophysical sources, $\mathcal{E}(\hat{\mathbf{s}}, f, t)$, in the sky coordinate system denoted by the unit vector $\hat{\mathbf{s}}$, propagate towards the observer as:
\begin{align}
  E(\mathbf{r}, f, t) &= \int E(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}\cdot\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, $\mathbf{r}$ denotes the observer's location, {\bf $f$ is the frequency of radiation, $c$ is the speed of light, $t$ denotes time} and $E(\mathbf{r}, f, t)$ is the propagated electric field. Thus the propagated electric field is a linear superposition of the electric fields emanating from astronomical sources with appropriate complex phases. {\bf Ignoring wide-field effects}, it can be {\bf simplified as} a Fourier transform of the electric fields in the sky coordinates.

An antenna, {\bf located at $\mathbf{r}_a$ (indexed by $a$)}, measures a phased sum of these propagated electric fields over its effective collecting area with an additive receiver noise:
\begin{align}\label{eqn:measured-E-field}
  E_a(f,t) &= \int W_a(\mathbf{r}-\mathbf{r}_a,f,t)\,E(\mathbf{r},f,t)\,\dif^2\mathbf{r} + n_a(f,t) \\
           &= \int W_a(\mathbf{r}-\mathbf{r}_a,f,t) \nonumber \\
           &\quad \times \left[ \int \mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}\cdot\hat{\mathbf{s}}/c}\,\dif\Omega \right] \dif^2\mathbf{r} + n_a(f,t) \\
           &= \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega + n_a(f,t)
\end{align}
where, {\bf $W_a(\mathbf{r},f,t)$ is the aperture electric field illumination pattern of the antenna and its Fourier transform, $\mathcal{W}_a(\hat{\mathbf{s}},f,t)$, is the directional antenna voltage response at a given frequency and time}.

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. {\bf The noise-like cross-terms ideally have zero mean}. This expression can be further simplified with the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)= \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{W}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power response, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}
V_{ab}(f,t) &= \int \mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by the multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  V_{ab}(f,t) &= \int W_{ab}(\mathbf{r}_{ab}^\prime-\mathbf{r},f,t) \nonumber\\ 
              &\qquad \times \left[\int I(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}.\hat{\mathbf{s}}/c}\,\dif\Omega\right]\dif^2\mathbf{r}\,+\, n_{ab}(f,t),
\end{align}
where, $W_{ab}(\mathbf{r},f,t)$ denotes the {\bf power response of the antenna pair obtained in the measurement plane} by a spatial Fourier transform of $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$. Effectively, the multiplication in image space by $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ has been replaced by a convolution with $W_{ab}(\mathbf{r},f,t)$ in {\bf measurement plane}. This is the software holographic equivalent of a traditional FX correlator output.

Hereafter, we adopt the matrix notation of \citet{mor11}, where vectors are represented with single coordinates, and matrices are represented by two coordinates denoting the spaces the operator transforms between. Since each frequency is processed independently, we drop the dependence on $f$. In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \mathbf{B}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ is Fourier transformed using $\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence function is weighted and summed using the antenna power response, $\mathbf{B}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \mathbf{E}(\mathbf{a})\,\mathbf{E}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating measured antenna electric fields over all possible pairs of $\mathbf{a}$ and $\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in matrix notation.

Using the optimal map-making formalism \citep{teg97a,teg97b}, a software holography image is formed using \citep{mor09}:
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{u})\,\mathbf{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v},\mathbf{v})\,\mathbf{m}(\mathbf{v}) \label{eqn:dirty-image-FX}
\end{align}
where the measured visibilities are weighted by the inverse of the system noise, followed by a gridding process using the holographic antenna power response as the gridding kernel, followed by a Fourier transform to create an image $\boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}})$. This is the optimal estimate of the true image $\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ given the visibility measurements.

The intermediate step of gridding with the antenna power response can be expressed as a convolution of a data vector generated by gridding the electric fields directly with the antenna illumination pattern.
\begin{align}
\mathbf{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v}) &= \left\langle \left[\mathbf{W}_a(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-\mathrm{T}}\!(\mathbf{a},\mathbf{a})\, \mathbf{E}^\star(\mathbf{a})\right]\right. \nonumber\\ 
&\quad\ast\left.\left[\mathbf{W}^\textrm{T}_a(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

We can then use the multiplication-convolution theorem to move the convolution in Equation \ref{eqn:e-field-conv} to a square after the Fourier transform in Equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\mathbf{N}^{-1}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

@@

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The propagated electric fields are shown on the left at different time stamps, $t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by antennas are denoted by $\widetilde{E}_1(t)\ldots \widetilde{E}_\textrm{N}(t)$. The F-engine performs a temporal Fourier transform on the electric field time-series to obtain electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\mathbf{E}(\mathbf{a})$ in matrix notation) for each of the antennas. Each of the complex antenna gains are calibrated to correct the corresponding electric field spectra. These calibrated electric fields are gridded using an antenna-based gridding convolution function after which it is spatially Fourier transformed and squared to obtain an image cube for every time stamp. These images are then time-averaged to obtain the accumulated image $\mathcal{I}^\prime(f)$ ($\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ in matrix notation).

\begin{figure}
  \includegraphics[width=\columnwidth]{figure1}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields shown on the left are measured as time-series $\widetilde{E}_1(t)\ldots \widetilde{E}_\textrm{N}(t)$ by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then Fourier transformed by the F-engine to produce electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series are imaged to produce images $\mathcal{I}^\prime_1(f)\ldots \mathcal{I}^\prime_\textrm{M}(f)$. These images are time-averaged to obtain the final image $\mathcal{I}^\prime(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility-based software holographic imaging from a FX correlator. The antenna-based F-engine is identical to that in the MOFF processing. The electric field spectra from each antenna are then cross-multiplied in the X-engine with those from all other antennas to obtain the visibilities $V_\textrm{ab}(f,t)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation). They are calibrated and time-averaged to obtain $\langle V_\textrm{ab}(f)\rangle$ which are then gridded and imaged to obtain the image $\mathcal{I}^\prime(f)$. The $\mathcal{I}^\prime(f)$ obtained from both techniques are theoretically identical as explained in \S\ref{sec:math}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure2}
  % \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility-based software holographic imaging in EPIC. The FX process flow shares the F-engine with the MOFF process. Following the F-engine, the electric fields pass through the X-engine to produce visibilities $V_\textrm{ab}(f,t)$ which are calibrated and time-averaged. Then they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which are then Fourier transformed to obtain the image, $\mathcal{I}^\prime(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

The gridding kernel in the simplest case is given by the antenna aperture
illumination function, $W(\mathbf{r}-\mathbf{r}_a)$, which can
be specified either by a functional form or as a table of values against 
locations around the antennas. A nearest neighbor mapping from all antenna 
footprints to grid locations is created using an efficient k-d tree algorithm 
\citep{man99}. There is no restriction here that the aperture illumination 
function has to be identical across antennas.

@@

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid, $W(\mathbf{r})$, is obtained by a projection of the individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  W(\mathbf{r}) &= \sum_a W_a(\mathbf{r}-\mathbf{r}_a),
\end{align}
or in matrix notation,
\begin{align}
  \mathbf{W}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathbf{I}(\mathbf{a}),
\end{align}
where, $\mathbf{I}(\mathbf{a})$ is a row of ones denoting coverage of the grid by kernel footprints of antennas. This is achieved by efficient multiplication with the sparse matrix created in the antenna-to-grid mapping process using the sparse matrices module in Python SciPy package. Unless $\mathbf{W}(\mathbf{r})$ includes time-dependent effects of the ionosphere or the instrument, it needs to be computed just once for the entire observation. However, the gridding of electric fields must be computed at every readout of the electric field spectra,
\begin{align}
  \mathbf{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathbf{E}(\mathbf{a}).
\end{align}

@@

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centred at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility-based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration time-scale and does not add significant cost to the full operation.

@@

Here we demonstrate our software using narrow band data from the LWA station 
in New Mexico. This data is in LWA narrow-band transient buffer (TBN) format 
from 255 antennas within roughly a diameter of 100~m. The data is centred at a 
frequency of 74.03~MHz, with a sample rate (equal to the bandwidth) of 100~kHz 
with 512 complex time samples per antenna in a A/D writeout time-scale of 
5.12~ms, a frequency resolution of 195.3125~Hz and dual polarization. There are 
391 such writeouts (or time stamps; each contains 512 time samples at 100~kHz 
sampling) yielding a total duration of 2~s.

@@

Besides its core functions of spatial Fourier transform and squaring, it can 
stack, accumulate and average images, and optionally remove the antenna
auto-correlations centred around the zero-spacing pixel in the $uv$ plane. 
It also handles all four cross-polarization products. Currently, it supports 
writing data out in standard FITS format.
